Carryover of a saddle-node bifurcation after transformation of a parameter into a variable

### 2020-09-16 Manuscript/article.tex
paper intro

Consider a biological system that has a saddle-node bifurcation. Saddle-node bifurcations are associated with meaningful properties such as biological switches and hysteresis. Would it be possible that a saddle-node bifurcation is also present if we transform one of the parameters to a variable? One would expect that such biological property in also present in an extended more complex version of the system. If yes, what is the relationship between the saddle-node bifurcations in the original and extended systems? Those bifurcations may not be associated meaning that the biological property is not shared between the two systems.

In the general case, let
\begin{equation}
    \dot x = f(x;\mu_{1},\mu_{2}),
    \label{equ:OriginalSystem}
\end{equation}
where $x\in\mathbb{R}^{n}$ and $\mu_{1},\mu_{2}\in\mathbb{R}$, be the \textit{original system} with a saddle-node bifurcation as one of the parameters crosses a critical value. Using continuation we can determine regions in the parameter space ($\mu_{1}\mu_{2}$-plane) where there are two more (less) steady states and the boundary where the bifurcation occurs. Now, transform one of the parameters, for example, $\mu_{1}$, into a variable to define the \textit{extended system}
\begin{equation}
    \begin{aligned}
    \dot x &= f(x,\mu_{1}; \mu_{2}), \\
    \dot \mu_{1} &= g(\mu_{1}; \mu_{2}),
    \end{aligned}
    \label{equ:ExtendedSystem}
\end{equation}
where $(x, \mu_{1})\in\mathbb{R}^{n+1}$ and $g(\mu_{1}; \mu_{2})$ is the vector field of the new variable $\mu_{1}$. We are interested in conditions for which a saddle-node bifurcation occurs as $\mu_{2}$ varies. From the bifurcation boundary in the $\mu_{1}\mu_{2}$-plane we can infer places in which a saddle-node bifurcation might occur as seen from the $\mu_{2}$-axis. When a saddle-node bifurcation does occur in the extended system \eqref{equ:ExtendedSystem} that is related to the saddle-node bifurcation in the original system \eqref{equ:OriginalSystem}, we call this the \textit{carryover of a saddle-node bifurcation} in the original system to the extended system.

Note that we can further generalize our study by considering multi-parametric spaces ($\mu \in \mathbb{R}^{m}$) and function $g$ depending on variables $z$. This is beyond the scope of this paper, but we briefly discuss this case in Section~\ref{sec:Discussion}.

Consider the dimensionless model for the activation of gen $x$ by biochemical substance $s$
\begin{equation}
    \dot x = s - rx + \frac{x^{2}}{1+x^{2}},
    \label{equ:Example:GenActivation:1D}
\end{equation}
where $r>0$ is the degradation rate and $s\geq0$ \citep{Strogatz1994, Lewis1977}. This model is characterized by an irreversible switch-like activation (critical transition) of gen $x$ when $s$ increases from zero above the threshold $s*=\frac{1}{2}$. This means that the activating biological signal needs to be high enough to activate the gen. Suppose that the degradation rate $r$ varies over time and that we incorporate its dynamics into the system, can we preserve the critical transition characteristic of gen $x$? and if yes, can we determine conditions for which the critical transition exists? The critical transition in \eqref{equ:Example:GenActivation:1D} is determined by a critical saddle-node bifurcation where the stable equilibrium in which gen activity is low collides with an unstable equilibrium, leading trajectories towards the remaining stable equilibrium in which gen activity is high. If $r$ is transformed into a variable, we would expect the same critical saddle-node bifurcation to occur but in a higher dimension. In that case, we say that we extend the smaller system and that we carryover the bifurcation to the extended system.

Consider also the regulatory network for cell division involving Cyclin B and APC (Anaphase Promoting Complex). High activity of APC determines cell division, but APC and Cyclin B regulate each other in a mutual antagonism keeping APC activity low. The active form of Cdc20, which increases as the cell grows, activates APC driving cell division. Is this regulatory network the cell mass acts an parameter driving the critical transition, i.e., there is . The model is given by
\begin{equation}
    \begin{aligned}
        \frac{dY}{dt} &= k_{1} - (k_{2p} + k_{2pp}P)Y, \\
        \frac{dP}{dt} &= \frac{(k_{3p}+k_{3pp}A)(1-P)}{J_{3}+(1-P)} - k_{4m}\frac{YP}{J_{4}+P}, \\
        \frac{dA}{dt} &= k_{5p} + k_{5pp}\frac{(mY/J_{5})^{n}}{1+(mY/J_{5})^{n}} - k_{6}A,
    \end{aligned}
    \label{equ:Example:CyclinActivation:1D}
\end{equation}
where $Y$ is the concentration of CyclinB, $P$ is the concentration of Cdh1 (part of APC), $A$ is the concentration of active of Cdc20, and $m$ is the cell mass \citep{Segel2013,Tyson2001}. We are interested in transforming the cell mass into a variable, yet the critical transition determining cell division must exist in the extended system.

@@

In this paper, we provide conditions in which the extended system \eqref{equ:1DCase:ExtendedSystem} has a saddle-node bifurcation provided the original system \eqref{equ:1DCase:OriginalSystem} has a saddle-node bifurcation. The we apply our findings to the gen activation \eqref{equ:Example:GenActivation:1D} and cell cycle division \eqref{equ:Example:CyclinActivation:1D} applications.

\section{Background}

@@

\section{Applications}

@@

\begin{figure}
    \begin{center}
    \includegraphics[width=0.48\linewidth]{Manuscript/figures/ApplicationGenPhasePortrait.pdf}
    \includegraphics[width=0.48\linewidth]{Manuscript/figures/ApplicationGenTwoParBifurcation2.pdf}
    \includegraphics[width=0.48\linewidth]{Manuscript/figures/ApplicationGenExtended2Bifurcation.pdf}
    \includegraphics[width=0.48\linewidth]{Manuscript/figures/ApplicationGenTwoParBifurcation1.pdf}
    \end{center}
    \caption{a) Phase plot for $s=0$ and $r=0.4$. The critical transition occurs when $s$ increases and the left and middle equilibria collide and disappear at $s^*\approx 0.0418$, forcing gen activity to the remaining stable equilibria on the right. b) Two-parameter bifurcation diagram. The enclosed region corresponds to bistability (as shown in a)). The cusp corresponds to $x=\sqrt{3}/3$ (nondegeneracy condition not met) where a cusp bifurcation occurs. The blue horizontal (vertical) line is the nullcline $g=0$ of the extended system. c) Bifurcation diagram for the extended system. [d) Two-parameter bifurcation diagram. Showing both cases (transforming $r$ or $s$), keep one of them after discussion]}
    \label{fig:Example:GenActivation}
\end{figure}

@@

Now, consider transforming $r$ into a variable with linear synthesis and degradation terms
\begin{equation}
    \begin{aligned}
        \dot x & = f(x,r;s) = s - rx + \frac{x^{2}}{1+x^{2}}, \\
        \dot r & = g(r;s) = a - br.
    \end{aligned}
    \label{equ:Example:GenActivation1:2D}
\end{equation}
where, $a>0$ and $b>0$.

>>>>

In biological systems, saddle-node bifurcations are associated with biologically meaningful properties such as biological switches and hysteresis. When adding extra dynamics to a biological system it might important that we preserve the existence of the bifurcation. Would it be possible to guarantee that a saddle-node bifurcation is also present if we add an equation? If yes, would those two bifurcation be associated to the same biological device? For example, if we are interested in modelling the dynamics of a parameter that drives the saddle-node bifurcation and turn into a variable, the bifurcation might still be present in the extended system with one extra dimension but with respect to another parameter. We call this property the \textit{carryover of a saddle-node bifurcation}.

In the general case, denote the \textit{original system}
\begin{equation}
    \dot z = f(z;\mu_{1},\mu_{2}),
    \label{equ:OriginalSystem}
\end{equation}
where $z\in\mathbb{R}^{n}$ and $\mu_{1},\mu_{2}\in\mathbb{R}$,  which has a saddle-node bifurcation as one of the parameters crosses a critical value. Using continuation we can determine the set of points in the parameter space where the bifurcation occurs. Now, transform one of the parameters, for example, $\mu_{1}$, into a variable to define the \textit{extended system}
\begin{equation}
    \begin{aligned}
    \dot z &= f(z,\mu_{1}; \mu_{2}), \\
    \dot \mu_{1} &= g(\mu_{1}; \mu_{2}),
    \end{aligned}
    \label{equ:ExtendedSystem}
\end{equation}
where $(z, \mu_{1})\in\mathbb{R}^{n+1}$ and $g(\mu_{1}; \mu_{2})$ is the vector field of the new variable $\mu_{1}$. We are interested in conditions for which a saddle-node bifurcation occurs as $\mu_{2}$ varies. If the bifurcation point in the extended system corresponds to a bifurcation point in the original system, then we can say that the bifurcation point of the former is the carryover of the latter.

Consider the following two examples. First, consider the dimensionless model for the activation of gen $x$ by biochemical substance $s$ given by
\begin{equation}
    \dot x = s - rx + \frac{x^{2}}{1+x^{2}},
    \label{equ:Example:GenActivation:1D}
\end{equation}
where $r>0$ is the degradation rate and $s\geq 0$ \citep{Strogatz1994, Lewis1977}. This model is characterized by an irreversible switch-like activation (critical transition) of gen $x$ when $s$ increases from zero above the threshold $s^*$. This means that the activating biological signal needs to be high enough to activate the gen. Suppose that we include the dynamics of substrate $s$ into the system such that the gen becomes always becomes active,
% Suppose that the degradation rate $r$ varies over time and that we incorporate its dynamics into the system, 
can the use the degradation rate in order to inactivate the gen?
% The critical transition in \eqref{equ:Example:GenActivation:1D} is determined by a critical saddle-node bifurcation where the stable equilibrium in which gen activity is low collides with an unstable equilibrium, leading trajectories towards the remaining stable equilibrium in which gen activity is high. If $s$ is transformed into a variable, we would expect a similar critical saddle-node bifurcation to occur but in a higher dimension.

Second, consider the regulatory network for cell division involving Cyclin B and APC (Anaphase Promoting Complex). High activity of APC determines cell division, but APC and Cyclin B regulate each other in a mutual antagonism keeping APC activity low. The active form of Cdc20, which increases as the cell grows, activates APC driving cell division. Is this regulatory network the cell mass acts a parameter driving the critical transition, i.e., there is a saddle-node bifurcation in which the activity of APC increases abruptly as the cell mass increases. The model is given by
\begin{equation}
    \begin{aligned}
        \frac{dY}{dt} &= k_{1} - (k_{2p} + k_{2pp}P)Y, \\
        \frac{dP}{dt} &= \frac{(k_{3p}+k_{3pp}A)(1-P)}{J_{3}+(1-P)} - k_{4m}\frac{YP}{J_{4}+P}, \\
        \frac{dA}{dt} &= k_{5p} + k_{5pp}\frac{(mY/J_{5})^{n}}{1+(mY/J_{5})^{n}} - k_{6}A,
    \end{aligned}
    \label{equ:Example:CyclinActivation:1D}
\end{equation}
where $Y$ is the concentration of CyclinB, $P$ is the concentration of Cdh1 (part of APC), $A$ is the concentration of active of Cdc20, and $m$ is the cell mass \citep{Segel2013,Tyson2001}. Supose that we are interested in transforming the cell mass into a variable so that cell division occurs for the appropriate initial conditions, is there another parameter that drives the cell division when the cell is unable to grow large enough?

@@

In both cases, we are interested in adding more dynamics to the model yet being able to preserve the biological characteristic defined by the saddle-node bifurcation. In this paper, we find conditions in the extended system \eqref{equ:ExtendedSystem} for the carryover of a saddle-node bifurcation in the original system 
\eqref{equ:OriginalSystem}. After reviewing the basic concepts of a saddle-node bifurcation in Section~\ref{sec:Background}, we first introduce our results for the one-dimensionas case in Section~\ref{sec:1DCase}, and then we extend out study to the $n$-dimensional case in Section~\ref{sec:NDCase}. We also provide a graphical and practical approch to guarantee the carryover of a saddle-node bifurcation. In Section~\ref{sec:Applications}, we apply our results to the two biological systems presented above. Note that we can further generalize our study by considering multi-parametric spaces ($\mu \in \mathbb{R}^{m}$) and function $g$ depending on $z$ in \eqref{equ:ExtendedSystem}. This is beyond the scope of this paper, but we briefly discuss this case in Section~\ref{sec:Discussion} along with further study.

% In this paper, we provide conditions in which the extended system \eqref{equ:1DCase:ExtendedSystem} has a saddle-node bifurcation provided the original system \eqref{equ:1DCase:OriginalSystem} has a saddle-node bifurcation. The we apply our findings to the gen activation \eqref{equ:Example:GenActivation:1D} and cell cycle division \eqref{equ:Example:CyclinActivation:1D} applications.

\section{Mathematical Background}
\label{sec:Background}

@@

\section{Applications}
\label{sec:Applications}

@@

\begin{figure}
    \begin{center}
    \includegraphics[width=0.98\linewidth]{figures/ApplicationGen.pdf}
%     \includegraphics[width=0.48\linewidth]{figures/ApplicationGenPhasePortrait.pdf}
%     \includegraphics[width=0.48\linewidth]{figures/ApplicationGenTwoParBifurcation2.pdf}
%     \includegraphics[width=0.48\linewidth]{figures/ApplicationGenExtended2Bifurcation.pdf}
%     \includegraphics[width=0.48\linewidth]{figures/ApplicationGenTwoParBifurcation1.pdf}
    \end{center}
    \caption{a) Phase plot for $s=0$ and $r=0.4$. The critical transition occurs when $s$ increases and the left and middle equilibria collide and disappear at $s^*\approx 0.0418$, forcing gen activity to the remaining stable equilibria on the right. b) Two-parameter bifurcation diagram. The enclosed region corresponds to bistability (as shown in a)). The cusp corresponds to $x=\sqrt{3}/3$ (nondegeneracy condition not met) where a cusp bifurcation occurs. The blue horizontal (vertical) line is the nullcline $g=0$ of the extended system. c) Bifurcation diagram for the extended system. [d) Two-parameter bifurcation diagram. Showing both cases (transforming $r$ or $s$), keep one of them after discussion]}
    \label{fig:Example:GenActivation}
\end{figure}

@@

Now consider transforming $r$ into a variable with linear synthesis and degradation terms
\begin{equation}
    \begin{aligned}
        \dot x & = f(x,r;s) = s - rx + \frac{x^{2}}{1+x^{2}}, \\
        \dot r & = g(r;s) = a - br.
    \end{aligned}
    \label{equ:Example:GenActivation1:2D}
\end{equation}
where, $a>0$ and $b>0$.
